---
title: "Untitled"
output:
  html_document: default
  word_document: default
---
```{r}
#make theme for all plots
fte_theme <- function(){
  color.background = 'ghostwhite'
  color.grid.major = 'black'
  color.axis.text = 'black'
  color.axis.title = 'black'
  color.title = 'black'
  theme_bw(base_size = 9) + 
    theme(panel.background = element_rect(fill=color.background,color = color.background)) +
    theme(plot.background = element_rect(fill = color.background, color = color.background)) +
    theme(panel.border = element_rect(colour = 'black')) +
    theme(panel.grid.major = element_blank()) + 
    theme(panel.grid.minor = element_blank()) + 
    theme(axis.ticks = element_blank()) +
    theme(plot.title = element_text(color = color.title, size = 15, vjust = 1.25)) +
    theme(axis.text.x = element_text(size = 8, color = color.axis.text, angle = 90)) + 
    theme(axis.text.y = element_text(size = 10, color = color.axis.text)) + 
    theme(axis.title.x = element_text(size = 12, color = color.axis.title, vjust = 0)) +
    theme(axis.title.y = element_text(size = 12, color = color.axis.title, vjust = 1.25)) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(axis.line.x = element_line(color="black", size = 0.15),
          axis.line.y = element_line(color="black", size = 0.15)) 
  
}
```


```{r}
#loading in data
predprey<-read.csv('predprey.csv')
library(tidyverse)

```

#Looking at effect of Latitude on predator and prey mass, using ggplot
```{r}
masslat_p<-ggplot(data=predprey, aes(x=Latitude, y=Prey_mass))+
  geom_point()+
  geom_smooth(method='lm')+
  fte_theme()
masslat_p+ xlab("Latitude")+ ylab("Prey mass (mg)")
masslat_P<-ggplot(data=predprey, aes(x=Latitude, y=Predator_mass))+
  geom_point()+
  geom_smooth(method='lm')+
  fte_theme()
masslat_P+ xlab("Latitude")+ ylab("Predator mass (mg)")

#Trendlines show majority of mass stays consistent throughout latitudes, with lots of variation around 40degrees (which was more sampled compared to the other latitudes)

```
Check to see if linear model for pred mass is significant
```{r}
predlong<- lm(predprey$Predator_mass~predprey$Longitude) #significant and negative 
summary(predlong)
plot(predlong) #using these to check normality of data, and lm assumptions. 

predlat<-lm(predprey$Predator_mass~predprey$Latitude) #significant and positive
summary(predlat)
plot(predlat)
```
```{r}
preylat<-lm(predprey$Prey_mass~predprey$Latitude) #significant and positive
summary(preylat)
plot(preylat)

preylong<-lm(predprey$Prey_mass~predprey$Longitude) #not significant
summary(preylong)
plot(preylong)
```


#Looking at how data is distributed between longitude~latitude
```{r}
longlat<- lm(predprey$Latitude~predprey$Longitude)
plot(longlat)

```
Very unregular, possibly pursue a log transform?

#Looking at taxonomic diversity in general and over space
    
```{r}
ggplot(data = predprey, aes(x = Predator_common_name)) +
  geom_bar(position = position_dodge(width=2))+
  fte_theme()
ggplot(data = predprey, aes(x = Prey_common_name)) +
  geom_bar(position = position_dodge(width=2))+
  fte_theme()
```
We can see that there are many species who only appear a few times (rare), while others form the majority of all data points.

Now let's plot species richness over latitude, separating out hemispheres:
```{r}
predprey2<-predprey %>% 
  mutate(Hemisphere =ifelse(Latitude>0, 'Northern_Hemisphere', 'Southern_Hemisphere')) #this part is dale's code but I need it for the below to work

predprey2 <- predprey2 %>% 
  mutate(Prey_Species_Richness=length(unique(Prey_common_name)), 
         Predator_Species_Richness=length(unique(Predator_common_name)))
predprey2N <- predprey2 %>% 
  filter(Hemisphere=="Northern_Hemisphere") #Separate out hemispheres to look more specifically. 
predprey2S <- predprey2 %>%
  filter(Hemisphere=="Southern_Hemisphere")

PredatorrichlatN <- predprey2N %>% 
  group_by(Latitude) %>% 
  summarize(Predator_Species_Richness=length(unique(Predator_common_name))) %>% #find the unique number of species at each latitude
  ggplot(.,aes(x=Latitude, y=Predator_Species_Richness))+
  geom_point()+
  geom_smooth(method = 'lm')+
  fte_theme()+
  labs(x = 'Latitude', y = 'Predator Species Richness')
PredatorrichlatN

PredatorrichlmN <- lm(Predator_Species_Richness ~ Latitude, data = predprey2N)
summary(PredatorrichlmN) #significant but wrong direction
plot(PredatorrichlmN)


PreyrichlatN <- predprey2N %>% 
  group_by(Latitude) %>% 
  summarize(Prey_Species_Richness=length(unique(Prey_common_name))) %>% 
  ggplot(.,aes(x=Latitude, y=Prey_Species_Richness))+
  geom_point()+
  geom_smooth(method = 'lm')+
  fte_theme()+
  labs(x = 'Latitude', y = 'Prey Species Richness')
PreyrichlatN

PreyrichlmN <- lm(Prey_Species_Richness ~ Latitude, data = predprey2N)
summary(PreyrichlmN) #significant but wrong direction
plot(PreyrichlmN)


PredatorrichlatS <- predprey2S %>% 
  group_by(Latitude) %>% 
  summarize(Predator_Species_Richness=length(unique(Predator_common_name))) %>% #find the unique number of species at each latitude
  ggplot(.,aes(x=Latitude, y=Predator_Species_Richness))+
  geom_point()+
  geom_smooth(method = 'lm')+
  fte_theme()+
  labs(x = 'Latitude', y = 'Predator Species Richness')
PredatorrichlatS

PredatorrichlmS <- lm(Predator_Species_Richness ~ Latitude, data = predprey2S)
summary(PredatorrichlmS) #not sig
plot(PredatorrichlmS)

PreyrichlatS <- predprey2S %>% 
  group_by(Latitude) %>% 
  summarize(Prey_Species_Richness=length(unique(Prey_common_name))) %>% 
  ggplot(.,aes(x=Latitude, y=Prey_Species_Richness))+
  geom_point()+
  geom_smooth(method = 'lm')+
  fte_theme()+
  labs(x = 'Latitude', y = 'Prey Species Richness')
PreyrichlatS

PreyrichlmS <- lm(Prey_Species_Richness ~ Latitude, data = predprey2S)
summary(PreyrichlmS) #not sig
plot(PreyrichlmS)

```



#PCA to see correlations between data
```{r}
library(car)
library(psych)
library(multcomp)

#select only the traits we think will have biological meaning
cutpredprey<-predprey %>% 
  dplyr::select(Predator_total_length, Predator_mass, Prey_length, Prey_mass, Latitude, Longitude, Depth, Mean_annual_temp)

#change all neccessary traits to be numeric, to do PCA
str(cutpredprey)
cutpredprey$Depth<- as.numeric(cutpredprey$Depth)

cutpredprey<- cutpredprey[complete.cases(cutpredprey),] #Only use complete cases to make things easier
predprey_cor <- cor(cutpredprey) #create a correlation matrix
det(predprey_cor) #we want to see that our determinant is greater than the necessary value of 0.00001, and it is
```
Running the first principal test:
```{r}
pc1 <- principal(predprey_cor, nfactors =8, rotate = "none") #Put 8 factors because this is how many traits we left in
pc1
#There are 3 SS loadings with a value over 1, meaning only 3 factors are probably important
```
```{r}
plot(pc1$values, type = "b") #3 components from the scree plot is also a reasonable estimate
```
Let's run the function again with only 3 factors:
```{r}
pc2 <- principal(predprey_cor, nfactors = 3, rotate = "none")
pc2
```
Finally, let's run the function using rotate to adjust the values:
```{r}
pc3 <- principal(predprey_cor, nfactors = 3, rotate = "varimax")
pc3
```
```{r}
a<-print.psych(pc3, cut = 0.3, sort = TRUE) #To create a nice read out of our groups

```
The groups formed here make some sense, and can inform members on what to test next. Prey length, prey mass and predator mass are a group. Mean annual temp, depth, and a negatively correlated longitude are a group. Longitude, latitude and predator total length also form a group.
